### 2017-04-05 paper/sexSelectionSpacing-ver1.tex
Further edits to theory section

@@

In the absence of prenatal sex determination, parents' only choice if they 
wanted a son was to try for another pregnancy as soon as possible after
the birth of a girl. 
This led to shorter spacing after the birth of a girl and what
is known as differential stopping behavior, where parents were 
more likely to cease child bearing after the birth of a son than
after a daughter [TK add references here]. 
As discussed above, the shorter spacing after the birth of a girl is 
associated with worse health outcomes for both mother and children.
Because of the potential health effects, most of the literature on spacing 
for countries like India naturally focus on whether spacing was too short.

With the introduction of sex-selective abortions, parents can abort
fetuses of unwanted sex.
With an approximately 12 months additional duration per abortion,
this strategy can result in very long durations between births, and
``too long'' spacing may become a greater concern for parents than 
``too short'' spacing.%
\footnote{
Shorter spacing was likely less of a concern to begin with for
better educated mothers---who are also the most likely to use 
sex selection---since their children are substantially less likely 
to be negatively affected by short spacing than children of less 
educated mothers \citep{Whitworth2002}.
}
If the duration from last birth becomes long enough parents may 
reverse their decision to use prenatal sex determination
and carry the next pregnancy to term whether it is a male or female fetus.

At least four different factors may make a reversal of the decision 
to use sex selection between one birth and the next more likely, 
everything else equal.
First, the more concerned parents are about possible infertility 
from repeated abortions or the closer they are to the end of their 
reproductive period, they more likely they are to decide that any 
child is better than not having an additional child at all.
Second, the higher the parents' discount factor is---that is, the
more the value present outcomes relative to future outcomes---the
more likely they are to reverse their decision to sex selection.
Third, economies of scale in childrearing---both in terms of time cost 
and direct cost---are easier to take advantage of if births 
are closer \citep[p 947]{Newman1984}.
Hence, the stronger these economies of scale are, the more likely
parents are to reverse their decision to use sex selection.
Finally, if women's wages increase with age, parents prefer
to have their children early and close together, while 
the opportunity cost of children is lower \citep{Heckman1976}.%
\footnote{
The time needed to care for young children can also impact 
timing if skills depreciate when out of the labor market \citep{Happel1984}.
Whether depreciation leads to longer or shorter spacing depends
on the specification of the depreciation function.
See also the discussion in \citet[p 315]{Hotz1997}
and references therein.
}
The more pronounced the increase in wage is with age, the more 
attractive it will be to reverse the decision to use sex selection.
Although, the latter two are most associated with the literature
on shorter spacing in middle- and  high-income countries, 
the increasing education and wealth in India means that a growing 
portion of the population is likely to behave in a similar manner.%
\footnote{
The possible exception is for the first birth, where the pressure to show 
fecundity---which shortens the duration to the first birth---is potentially 
an important factor in India \citep{dyson83,Sethuraman2007,Dommaraju2009}.
}

If ``too long'' spacing leads parents to reverse their decision to use
prenatal sex determination within a spell this has important implications 
for the empirical method.

@@

Lowering the cost of prenatal sex determination does, however, bias the overall sex 
ratio further towards boys, and the stronger the son preference and the 
more patient the parents, the larger the change in sex ratio.
Correspondingly, both the number of abortions per woman and the
average number of spells per woman where a prenatal scan is used also increases.
Since fertility changes little, the increased use of abortion
is associated with longer spacing between births.
This increase in spacing is, however, not uniform across parities.
In addition to increasing the overall sex ratio, lower cost of prenatal 
scan leads to use of sex selection at earlier parities.
The result is that in some cases there is either no change or a \emph{decrease} 
in average spacing for a given parity---although with a more than compensating 
increase in an earlier parity.
Hence, it is possible for individual parities to show a \emph{decline}
in sex selection at the same time as the overall use goes up.
This is an important caution against over-interpreting
individual parity results.
It also highlights the need to incorporate fertility progression into
estimates, something the simple model is incapable of.

% Son preference

Increasing the value of having sons, $\alpha$, when there is access to 
sex selection results in slightly lower fertility in most cases,
but the effect is neither strong nor uniform.
What is clear is that the sex ratio becomes more male biased 
the stronger the son preference.
Correspondingly, there is an increased number of abortions,
more spells where a prenatal scan is used,
and the spacing between births is longer.

Son preference, however, manifests in two ways in this model.
First, if $\alpha$ is larger than 0.5 parents gain more utility
from having an additional son than from an additional daughter---provided
the had the same number of sons and daughters before.
Second, parents will never abort a male fetus, only a female one.
This explains why even when parents have equal utility from an additional
son or daughter sex ratios still increases as cost of children 
increases and cost of prenatal scans decline.

At $\alpha = 0.5$ parents prefer an equal number of boys and girls.
The more children you have the less likely it is that you will
resort to sex selection to ensure a balance between sons and daughters.
First, the chances of having, for example, only boys or only
girls, decreases with each additional child.%
\footnote{
For 2 children the likelihood of only boys or only girls
is 50\%, but for 4 children it is 12.5\% (assuming equal probability
of a boy or a girl).
}
Second, the more children you have the lower the marginal
utility of having another, which means that, for a given price
of prenatal scan, the less attractive it is to bear the cost of using
prenatal scans to ensure a specific sex.
As fertility fall to, say, two, however, those parents who had a girl as their 
first child will now want to use prenatal sex determination to 
ensure a balanced number of sons and daughters and this is why
the sex ratio is so unbalanced as the cost of children increases.
Parents who had a son as their first child would also want to
use sex selection to ensure both a son \emph{and} a daughter, but since
that option is not available here they have their
second birth without scan.

Stretching the interpretation of the model, think of $\alpha=0.5$ 
as capturing parents treating sons and daughters equally once born.
The use of sex selection to achieve balance---but only when the
fetus is female---could be a potential explanation for the
suggestion that sons and daughters are treated relatively equally and 
we are still seeing an unequal sex ratio.
What is relevant here it that there is no way to identify which 
of these two different types of son preference affect sex ratios 
from fertility data.
This is a concern because the way we currently measure son preference
allows us to capture only the distribution preference, and 
the different types of son preference likely have different policy 
implications.

@@

Another reason for parents to change their decision to use sex
selection is that they get ``too close'' to the end of the time
horizon after extensive use of abortions and prefer any child to 
the potential of having no child at all.
This type of reversal explains why the effects of changing the discount 
rate and cost of prenatal sex determination are not uniform.
Related, extending the time horizon leads to a smaller likelihood of parents 
reversing their use of sex selection between births when
there is strong incentive to use sex selection.%
\footnote{
These results are available in the online Appendix.
}

>>>>

[ways to get at son preference:
- stated fertility preferences
- longer spacing after son
- differential stopping behavior
- biased sex ratio (overall before sex selection, and specifically at birth later)]

[why do these work less well after introduction of sex selection?]

[how can my method remedy those problems?]

[main results]

@@

In the absence of prenatal sex determination, parents' only choice if they 
wanted a son was to try for another pregnancy as soon as possible after
the birth of a girl. 
This led to shorter spacing after the birth of a girl and what
is known as differential stopping behavior, where parents were 
more likely to cease childbearing after the birth of a son than
after a daughter [TK add references here]. 
As discussed above, the shorter spacing after the birth of a girl is 
associated with worse health outcomes for both mother and children.
Because of the potential health effects, most of the literature on spacing 
for countries like India naturally focus on whether spacing was too short
and used the shorter birth spacing after the birth of a girl than after the
birth of a boy as a measure of the degree of son preference.

With the introduction of sex-selective abortions, parents can now abort
a fetus of unwanted sex.
With an approximately 12 months additional duration per abortion,
this can result in substantially longer durations between births, and
``too long'' spacing may become a greater concern for parents than 
``too short'' spacing.%
\footnote{
Shorter spacing was likely less of a concern to begin with for
better educated mothers---who are also the most likely to use 
sex selection---since their children are substantially less likely 
to be negatively affected by short spacing than children of less 
educated mothers \citep{Whitworth2002}.
}
If the duration from last birth becomes long enough, parents may 
reverse their decision to use prenatal sex determination
and carry the next pregnancy to term whether male or female.

A number of factors make a reversal of the decision to use sex 
selection between one birth and the next more likely, 
everything else equal.
First, the more concerned parents are about possible infertility 
from repeated abortions or the closer they are to the end of their 
reproductive period, they more likely they are to decide that any 
child is better than not having an additional child at all.
Second, the higher the parents' discount factor is---that is, the
more the value present outcomes relative to future outcomes---the
more likely they are to reverse their decision to sex selection.
Third, economies of scale in childrearing---both in terms of time cost 
and direct cost---are easier to take advantage of if births 
are closer \citep[p 947]{Newman1984}.
Hence, the stronger these economies of scale are, the more likely
parents are to reverse their decision to use sex selection.
Finally, if women's wages increase with age, parents prefer
to have their children early and close together while 
the opportunity cost of children is lower \citep{Heckman1976}.%
\footnote{
The time needed to care for young children can also impact 
timing if skills depreciate when out of the labor market \citep{Happel1984}.
Whether depreciation leads to longer or shorter spacing depends
on the specification of the depreciation function.
See also the discussion in \citet[p 315]{Hotz1997}
and references therein.
}
The more pronounced the increase in wage is with age, the more 
attractive it will be to reverse the decision to use sex selection.
Although the latter two factor are most often associated with the literature
on shorter spacing in middle- and  high-income countries, 
the increasing education and wealth in India means that a growing 
portion of the population is likely to behave in a similar manner.%
\footnote{
The possible exception is for the first birth, where the pressure to show 
fecundity---which shortens the duration to the first birth---is potentially 
an important factor in India \citep{dyson83,Sethuraman2007,Dommaraju2009}.
}

For a given sex composition of prior children, there are two main drivers 
of sex selection: the degree of son preference and the cost of sex selection.
The more parents care about having a son as their next child, the more
likely they are to use sex selection and the less likely they are to
reverse the decision once started.
The result is an increased number of abortions, more spells where a
prenatal scan is used, and the spacing between births is longer.

Lowering the cost of sex selection bias the overall sex ratio further 
towards boys.
Correspondingly, both the number of abortions per woman and the
number of spells where a prenatal scan is used should also increase,
resulting in longer average spacing between births.
This increase in spacing is, however, not likely to be uniform across parities.
In addition to increasing the overall sex ratio, lower cost of prenatal 
scan should also lead to use of sex selection at earlier parities.
It is therefore possible to see either no change or a \emph{decrease} 
in average spacing for a later parity---although with a more than compensating 
increase in an earlier parity.
In other words, it is possible for individual parities to show a \emph{decline}
in sex selection at the same time as the overall use goes up.
This is an important caution against over-interpreting individual 
parity results.

It also highlights the need to incorporate fertility progression into
estimates, something the simple model is incapable of.

If ``too long'' spacing leads parents to reverse their decision to use
prenatal sex determination within a spell this has important implications 
for the empirical method [TK explain what they are instead!].

@@

@@
